Whitepaper: Contract Modules

### 2020-12-04 papers/contract-modules/modules.tex
update

Here is a contract type that describes / specifies the above Michelson code.
\begin{lstlisting}[caption={Contract module example},label={lst:contract-module-example}]
contract type MyContract = sig
  paid entrypoint %bid_for_auction (value : int) 
  raises "closed"                 (** auction closed*)
       | "bid too low" as too_low (** bid received is too low *)
       | 42 as the_answer         (** failwith received argument 42 *)
       | (_ : int) as failedwith  (** failwith received an integer arg *)

@@

The body of the \lstinline/sig ... end/ can be stored in a
\texttt{.tzm} file where the basename matches the one of the
\texttt{.tz} file. The output is a \texttt{.mli} file which contains a
valid OCaml signature. In the next step, we are also generating the
corresponding \texttt{.ml} file implementing the signature.

Each \lstinline/entrypoint/ is mapped to a function. The final pattern
variable may be a wildcard \lstinline/_/ or an OCaml identifier, which
is used in generating the implementation. For \lstinline/_/ a standard
name will be generated.

@@

  val bid_for_auction
    : Tezos.pukh -> ?fee:Tezos.tz -> Tezos.tz
    -> int -> (Tezos.status, bid_for_auction_errors) Tezos.monad

@@

(* variable part *)
let contract_addr = 
  get_contract "KT1ThEdxfUcWUwqsdergy..." >>=? fun contract -> ???

@@

let big_for_auction_errors_map =
  let open Base in
  Base.Map.of_alist_exn (module String)
    [
      (failwith_regex ^ "closed", closed);
      (failwith_regex ^ "bid too low", too_low);
      (failwith_regex ^ "42", the_answer);
      (failwith_regex ^ int_regex, failedwith (0)); (* badness 10000 *)
    ]
  (* inadequate in several ways;
   * should just generate code instead of using a map*)

let bid_for_auction destination ?(fee=100.0) amount value =
  ...
  begin
    SyncAPIV0.call_contract
      amount
      destination
      contract_addr
      ~(serialize Tint value)
      fee
    >>=? fun oph ->
    wait_for_inclusion oph (* busy waiting *)
    >>=? fun _ ->
    return 0
  end
  >>= function
  | Ok _ -> ...
  | Error (Rejection Michelson_runtime_error s) -> ...

@@

let close_auction destination ?(fee=50.0) ?(amount=0.0) _ =
  ...
  SyncAPIV0.call_contract
    amount
    destination
    contract_addr
    ~(serialize Tunit)
    fee
\end{lstlisting}

@@

An extended type checker collects information about explicit failures along execution paths
starting from each possible entry point. This checker will be able to
verify the contract module specification in
\ref{lst:contract-module-example} against the Michelson code in
\ref{lst:code-example}. We may also wish to provide an entry point so
that the checker only gathers information for that entry point. This alternative may be interesting if there
is an entry point taking a parameter of sum type, rather than using all injections to distinguish
between entry points.

@@

>>>>

Here is a \textbf{contract module} that describes / specifies the above Michelson code.
\begin{lstlisting}[caption={Contract module example},label={lst:contract-module-example}]
contract type MyContract = sig
  paid entrypoint %bid_for_auction (value : int) 
  raises "closed"                 (** auction closed*)
       | "bid too low" as too_low (** bid received is too low *)
       | 42 as the_answer         (** failwith received argument 42 *)
       | (_ : int) as failedwith  (** failwith received an integer arg *)

@@

To specify the contract module, we store
the body of the \lstinline/sig ... end/ in a
\texttt{.tzm} file where the basename matches the one of the
\texttt{.tz} file which implements the contract.

A contract module gets compiled to an \texttt{.mli} file which contains a
valid OCaml signature and the corresponding \texttt{.ml} file
implementing the signature.

Each \lstinline/entrypoint/ is mapped to a function. The variables
used in the pattern may be a wildcard \lstinline/_/ or an OCaml identifier, which
is used in generating the implementation. For \lstinline/_/ a standard
name will be generated.

@@

  val bid_for_auction
    : Tezos.pukh -> ?fee:Tezos.tz -> amount:Tezos.tz
    -> int -> (Tezos.status, bid_for_auction_errors) Tezos.monad

@@

(* variable part *)
let contract_addr = 
  get_contract "KT1ThEdxfUcWUwqsdergy..." >>=? 
  fun ca -> ???

@@

let big_for_auction_errors_map =
  let open Base in
  Base.Map.of_alist_exn (module String)
    [
      (failwith_regex ^ "closed", closed);
      (failwith_regex ^ "bid too low", too_low);
      (failwith_regex ^ "42", the_answer);
      (failwith_regex ^ int_regex, failedwith (0)); 
      (* badness 10000: actual implementation must obtain the int value *)
    ]
  (* inadequate in several ways;
   * should just generate code instead of using a map*)

let bid_for_auction destination ?(fee=100.0) amount v =
  ...
  begin
    SyncAPIV0.call_contract
      amount
      destination
      contract_addr
      ~(serialize Tint v)
      fee
    >>=? fun oph ->
    wait_for_inclusion oph (* busy waiting *)
    >>=? fun _ ->
    return 0
  end
  >>= function
  | Ok _ -> ...
  | Error (Rejection Michelson_runtime_error s) -> 
    (*match s against expected errors from ...errors_map *)
    (*throw exception if no match *)
    ...

@@

let close_auction destination ?(fee=50.0) ?(amount=0.0) _ =
  ...
  SyncAPIV0.call_contract
    amount
    destination
    contract_addr
    ~(serialize Tunit)
    fee
\end{lstlisting}
This code example is incomplete and likely incorrect. It contains
fragments from an example implemented on top of the synchronous API.
The call to \lstinline/serialize/ refers to a proposal to improve on
the typing properties of the synchronous API.

@@

An extended type checker collects information about explicit failures
along execution paths starting from each possible entry point. This
checker will be able to verify the contract module specification in
Listing~\ref{lst:contract-module-example} against the Michelson code
in Listing~\ref{lst:code-example}. We may also wish to provide an
entry point so that the checker only gathers information for that
entry point. This alternative may be interesting if there is an entry
point taking a parameter of sum type, rather than using all injections
to distinguish between entry points.

@@

\subsection{The Checker}
\label{sec:checker}

\subsubsection{Input}
\begin{itemize}
\item A contract in the form of a \texttt{.tz} file containing a
  Michelson program.
\item An entrypoint in the program
\end{itemize}

The checker performs symbolic execution of the Michelson stack
machine.

A symbolic value is either a concrete value
constructor (like a number, a pair, left, right, nil, cons, etc) where
the components are symbolic values, or it is a Michelson type which
may be associated with a constraint, e.g., for an integer $x$ there 
may be the constraint that $0<x<10$. Subcomponents of the type are
again types with constraints. In a first iteration, constraints may be
restricted to a disjunction of equations, e.g., $x=1 \vee x=2$.

\subsubsection{Type checking}
\label{sec:type-checking}

 The stack is initialized with a symbolic value corresponding to a pair
of the storage type and the parameter type. The latter is concretized
according to the path to the specified entrypoint. That is, to
investigate the entrypoint \lstinline/bid_for_auction/ in
Listing~\ref{lst:code-example}, symbolic execution would start with a
one-element stack that contains
\begin{lstlisting}[language=michelson]
(PAIR (LEFT x1:int) x2:address)
\end{lstlisting}
without any constraints on \texttt{x1} or \texttt{x2}.

The checker executes instructions as far as possible using the
concrete part and updates the stack according to the typing rules and
semantics of the instruction. It tries to keep the
stack contents as concrete as possible by exploiting all known information
(e.g., constraints). If control flow cannot be decided, the checker
explores all alternatives, potentially adding constraints that reify the
checked condition, and joins the results. Execution paths with
unsatisfiable constraints are pruned.

\subsubsection{Exception tracking}
\label{sec:exception-tracking}

If the checker encounters an instruction that can make the contract
fail, it saves this potential failure along with the (control flow) constraints
leading to it. For a \texttt{FAILWITH} instruction, it saves
this failure along with the constraints and with the symbolic value on
top of the stack.

\subsubsection{Token tracking}
\label{sec:token-tracking}

Instructions that generate tokens (e.g., \texttt{AMOUNT},
\texttt{BALANCE}) create symbolic values marked with 
the instruction. Further processing of these symbolic values leads to
constraints on tokens and to their usage. These are also saved to
infer if an amount argument is needed.
